Weighing Chances and Use Against Risks and Misuse of Emerging Genome Analysis and Manipulation Technology

### 2022-05-31 draft.tex
genetic privacy

\subsection{Genetic Engineering}
Human interference in evolution to bend nature to our will goes back at least to 2000 B.C. with strong evidence that the Assyrians artificially pollinated date trees\cite[p.633]{mayr}.
Since then, several methods with the shared goal of reinforcing desirable characteristics and eradicating undesirables ones developed, most dominantly selective breeding, with the recent modification called mutation breeding, the practice of directly exposing organisms, usually seeds, to mutagenic material to positively augment the mutation rate and such the chance of new positive traits being present\cite{mutationbreeding}.
It can be argued that since selective and mutation breeding only ever affect coming generations it can hardly be seen as genetic modification of a single organism and is better labeled as a separate method of eugenics.
From that perspective the first milestone of genetic engineering was the Avery–MacLeod–McCarty experiment in 1944, which proved that DNA was responsible for transformation in virus strains. It built upon Griffith's experiment from 1928, which showed that potent but dead virus strains can still transfer their properties to usually harmless strains, via something called the "transforming principle".
Avery, MacLeod and McCarty proved that the relevant component for this to work was DNA and not like previously assumed specific proteins, since the phenomenon persist when the solution is treated with protein dissolving substance, but seizes to persist when treated with DNA dissolving agents.
The most direct and controlled form of genetic modification is inserting strings of DNA into existing cells to transfer properties from one organism to another, deactivating- or cutting them out to suppress traits.
The most prominent example of this to date is the CRISPR/Cas method, which accomplishes this semi-reliably via the use of bacteria, but currently still struggles with off target effects, i.e. unforeseen changes outside of the intended locality.
While originally discovered in 1987, its use for genetic editing was only demonstrated in 2011 and since then a tremendous amount of research went into improving effectivity, safety and specificity, as well as finding applications in numerous in agriculture and medicine.
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non-modified cells\cite{crispr}.
This can be worked around in three ways.
Modify a sufficient number of cells (most likely infeasible for broad changes), modify in an early stage of development, where only a number of cells exist, or germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

Secondly gene association studies, which aim to map genetic variants to observable traits and come in two varieties. 
Genome-wide association studies (GWAS) hope to uncover any correlation between strings of DNA and phenotype, while specific genetic association or gene linkage studies focus on specific variants of a gene (alleles) and examine whether they frequently cooccur with disease\cite{gwas}.
This endeavour presents numerous challenges stemming from the sheer complexity of the mechanisms at hand.
Part of the reason this differs from an exact science is changes an organisms environment has on its genome. It is estimated as much 98\% of human DNA is noncoding, id est lacks any apparent purpose\cite{noncoding}. Some of it however is not entirely dysfunctional, rather it is dynamically activated and silenced in a process known as epigenetics changes by so called regulatory genes\cite{epigenetics}.
This means the exact same sequence of DNA can express itself differently, making direct conclusions about the significance of a specific snippet fall short of reliability.
In 2003 with the official completion of the Human Genome Project (HGP) this has been accomplished\cite{HGPFAQ}.
The human genome consists of approximately 3.3 billion base pairs, by cross referencing a healthy genome, single point mutations can be effortlessly identified in theory, especially with modern high-throughput analysis tools\cite{rawal}.
Unfortunately in the context of genetics this notion of a healthy genome is vague. Not only does a healthy genome not exist, simply averaging the genome of a few relatively healthy adults would do it no justice. Changes in DNA can also come in the form of deletion or addition of bases rendering traditional similarity metrics useless as an applicable method would have to be alignment free, in other words agnostic to the exact position in sequence, rather it should recognize subsequences or repetitions anywhere they show up and compare the frequency. Through the nature of mutation, the 'average genome' also varies widely regionally, the HGP in particular faced criticism for overrepresenting caucasian DNA, owed to the fact that it was conducted on anonymous donations of DNA in the USA\cite{HGPcritic}.
It should therefore be seen more like a reference frame to possibly detect deviation.
Cataloguing mutations is useful but by itself often provides little insight, as chances are the mutation occurred a noncoding part.
Even when mutation occurs in regulatory sequences or as part of a gene single mutations these usually have little effect.
Its important to keep in mind that with mutations being so common, each cell has a slightly different sequence. The only mutations they share are those that occurred in the germ cells that formed the zygote or in the very early stages afterwards.
Germline mutations occur roughly $\frac{number of base pairs}{2}*10^{-9}$ times per year. In humans that works out to about 1.65 inheritable mutations per year per parent, but when examining the genome of adult humans there is sampling bias at play. Single point mutations with drastic changes usually just render the cells nonviable, consequently an egg being fertilized and developing long enough to be given birth to indicates a mostly intact genome\cite{scally}.
Somatic mutations occur in the order of magnitude of $ 2*\frac{number of cell divisions}{10^8 base pairs}$, which with 30 trillion cells equates to roughly 500 trillion per year\cite{somaticmutation}.
These usually have no or very little effect, are repaired by the cells dna repair mechanisms or kill the cell either by inhibiting vital function or by causing cell suicide\cite{dnarepair}.
The unlikely unlikely case that DNA damage results in neither of these scenarios and the cell instead starts rapidly reproducing is what we call cancer.

@@

Every individual has a unique genetic makeup, their own distinct form of the human genome. 
Furthermore, our DNA remains the same from the first instant of an individual’s existence to his or her last breath.\cite{dnaidentity} 
Sociologist Abels describes identity as the answer to the question \textit{who am I} and therefore, identity as a differentiating term.\cite{abelsidentity} Consequently, it is not unjustified to combine both. On the one hand, there is a distinctive combination of DNA sequences that is individual for everyone and consistent for life. 
On the other hand, there is a need for clarification about one's own identity, the will to differentiate oneself from others while remaining consistent within oneself. But what do people hope to gain from their genetic test results?

According to Nordgren, new information can lead to a new self-understanding.\cite{nordgren} The results of genetic testing allow an insight into irrefutable truths, when executed correctly by qualified personnel. 
In the case of a paternity test, a good example for this would be "I am P’s father". The statement is added as a new aspect of one's own perception and thus as part of one's own identity. Same procedure for the statement "I am an African-American with my roots in Nigeria", based on genetic ancestry tracing.

In a study published in 2018, about 30\% of respondents said they had taken a genetic test to learn how their DNA influences their physical or emotional traits.\cite{canceropinion} 
69 \% of Americans surveyed are very confident that there is a complex genetic code inside our cells that helps determine "who we are".\cite{apgfk} 
Apparently, the link between DNA results and information about one's own characteristics thus seems to have already reached some people. Whether the link was clear from the start or created through targeted marketing, it is probably a combination of both.

@@

DNA tests are more popular than ever. Due to the decline in the cost of genetic analysis in the last two decades, "direct-to-consumer" genetic screening has become affordable for more and more people\cite{rodney}. 
In 2018, the global direct-to-consumer market was valued at approximately 824.1 million USD and is estimated to grow to 6,364.5 million USD by 2028. 
The rise in the public awareness along with the reduced cost and time required for sequencing owing to the technological advancements is responsible for bolstering the industry demand and aiding in reduction of the global healthcare expenditure.\cite{bisresearch}

\begin{quote}
"Get insights from your DNA, whether it's your ethnicity or personal traits." (Ancestry.com)
\end{quote}

\begin{quote}
"Find out your likelihood of having certain characteristics. See how your DNA affects your hair color, taste preferences and more." (23andme.com)
\end{quote}

Many genetic testing companies appeal in their rhetoric to identity. The previous quotes use words like "characteristics" and "personal traits" to manifest the connection between DNA testing and the customers identity. 
Nordgreen describes this as genetic individualism which is the view that the DNA of an individual gives the individual a unique identity\cite{nordgren}.
This is in accordance with the statements from the previous chapter about DNA test results leading to a new self-understanding.
Furthermore, Nordgren state that modern societies are searching for identity and this rhetoric is a response to that\cite{nordgren}.
In addition to that, Nordgren and Juengst believe that the business models of these companies represent the confluence of three very different currents within contemporary Euro-American culture: the distinctly pre-modern search for a naturalistic understanding of individual identity in a pluralistic world, the thoroughly modern cachet of genomics as a science, and the post-modern emphasis on radical individual self-determination.\cite{nordgrenjuengst} The marketing methods of the websites partly explain the growing popularity of DNA tests, but what are the opportunities and risks?

@@

The two biggest market players Ancestry and 23andMe have a combined DNA Database Size of 32+ million\cite{ancestrydatabase}\cite{about23andme}. With this amount of data and due to the sensitivity of the content, it should be handled very responsibly. But how does it look in practice?
Laestadius, Rich \& Auer investigated 30 different direct-to-consumer genetic testing websites for two aspects. 
The first aspect being transparency regarding confidentiality and privacy. As a second aspect, the use of data, professional and government bodies created guidelines to promote transparency among these companies was analysed. 
The authors concluded that although most companies met guidelines related to transparency regarding security protocols, storage procedures, and third-party disclosures, only few met guidelines regarding sharing risks from data disclosures. 
Additionally, few companies disclosed how long data would be kept for services or research. 
Use of data for research was frequently mentioned only in privacy policies and terms of service documents, and only two-thirds of companies required an additional consent to use consumer data for health-related research. 
This analysis shows that companies do not consistently meet international transparency guidelines related to confidentiality, privacy, and secondary use of data and are therefore a risk for their customers. 
The authors appeal to clearly inform consumers about specific third parties that will be given access to their data, whether they can have their samples and data deleted at will, and whether their data will be used for anything other than the services they purchased.\cite{companydata}

@@

The controversy around privacy versus surveillance applies to genome analysis as much as to any sensitive technology. 
Germanys BKA describes the evolution of the usage of DNA analysis as going from a supplement to the "classic" identification method of fingerprints to an equal and self-evident means of securing evidence\cite{bkapres}. 
A common claim of advocates for DNA collection is that there is no difference between DNA and fingerprinting in terms of its invasion of an individual's privacy. 
But while fingerprints are a unique identifier, DNA identifies networks of people.\cite{cwduster} DNA data can reveal information about third parties that they have not consented to.
 In the case of the Golden state killer, Criminal DNA databases produced no hits but a match to a distant relative on a genealogy website has led to the identity and therefore the arrest of the perpetrator\cite{goldenstate}. 
In this incident, the misuse of DNA profiles on genealogy websites has led to the prevention of further crimes and therefore has added value for society.

@@

\subsection{Paternity}
After testing for genetic relatedness, non-legal and legal paternity tests are offered most often by direct-to-consumer genetic companies.\cite{dnastatista}
Genetic tests can be helpful in establishing evidence for the parenthood of a person for a case like child custody and support. 
The results of genetic test can also be used as a support for placing a parent’s name on the birth certificate of a child or for settling disputes in child custody laws. However, whether the sampling is done with the consent of all parties tested is not the responsibility of the genetic websites.

@@

\subsection{Medical Sector}
As preventing disease is usually preferable over treating it, a substantial amount of medicine focuses on early detection and prevention. Especially for the class of mendelian disorders, which are characterized by being linkable to a single gene defect, genetic testing has proven to make diagnosis easy and reliable \cite{rawal}.
However while a grand number of such diseases exist, they are usually rare so testing might only pay off if the defect is known to run in the family \cite{publicopinion}.
Much bigger potential lies in preventing common disease with either a genetic component or whose causes are spread over several genes and are therefore  harder to pinpoint. This uncertainty neccessarily introduces stochastics and the method known as genetic risk assessment \cite{gwas}.
Advances here have sprouted the field of pharmagenomics, which utilizes genetic information to practice precision medicine.
Just like ethnicity, genetic profile has been shown to affect drug response, thus using this knowledge to personalize prescriptions can increase effectiveness and lower cost and side effects \cite{costsaving}.
Predictive genetic testing does not come without concerns from both experts and the public, the most prominent ones will be discussed in the following:

\subsubsection{Patient Autonomy}
The primary role of any new medical instrument should be to serve, therefore patient autonomy and informed consent are non-negotiable, which necessitates  public education.
Broad, as well as targeted at populations most susceptible to specific genetic conditions.
Equally as important is access to testing facilities themselves as well as short wait times for both test and treatment\cite{publicopinion}.
\subsubsection{Genetic Discrimation}
As will be discussed in the following chapter, subpar treatment grounded in genome analysis results is no better than discrimination on the basis of sex, race, age or religion.
\subsubsection{Commercial Use}
Profit as the primary incentive manages to corrupt even the most honourable endeavour.
Commercial gene testing and therapy raises the same questions about motive and ethical obligation to help that plagued the medical ethics for millennia.
Testing costs being inflated because of patents can hardly be in accordance with public interest, which sparks discussion about the next point\cite{publicopinion}.
\subsubsection{Use of Public Resources}
If private patents are to be avoided, public resources have to be used towards research in the applicable area.
Distributing those in a fair way however proves difficult. Developing tests is expensive so some conditions need to be prioritized, which in turn requires weighing of different factors. The initial cost of development is just as important as benefit it brings. This is comprised of the reliability of the result, as well as therapy options versus their respective cost. An unreliable result is equally useless from a public health perspective as knowledge about a genetic death sentence.
Additionally as many tests would only ever affect a small number of people, some argue that the money should either be used for more general genetic tests or towards programs that target public health in a broader sense like addressing diet\cite{publicopinion}.
\subsubsection{Patient Responsibilities}
While a general right to data privacy in medicine makes sense for traditional diagnostics, one might argue that since DNA test results have immediate consequences for relatives theres at least in part a responsibility to inform said relatives of any risks they themselves might face, comparable to infectious disease or STDs\cite{publicopinion}.
This apparent obligation to warn stands in direct conflict with the testers pledge of silence and the testees right to privacy, a conflict that could in theory be resolved with institutions for anonymous testing and subsequent information of concerned parties analogous to HIV testing.
\subsubsection{Psychological Risks}
The psychological weight associated with carrying near certain information about the future is not to be underestimated in the context of genetic testing either.
Just as a cancer diagnosis, discovering to be at high risk of developing potentially lethal disease can be devastating, especially if treatment options are limited\cite{publicopinion}.
Depending on the evolution of public debate on the topic, a scenario in which gene therapy is frowned upon in a similar manner to termination of pregnancy by groups within society is also imaginable.
This would add additional stress and force people at risk for genetic disease to make a choice between social acceptability and their own health.

\subsection{Insurance}
Insurance as a business walks a fine line between solidarity and preying on the vulnerable. Risk and how we manage it as a society is studied under the name of insurance ethics, but to date the literature is sparse.
While public insurance in practice is a form of wealth redistribution, private insurance is in essence paying for peace of mind\cite{insuranceethics}.
Looking at the probability theory behind it, we expect to pay more than the 
However this social safety net only functions as long as there is uncertainty, in fact it functions better the more uncertainty there is.
This becomes obvious in consideration of the idea that clarity about the future replaces the insurance sector.
In a fully deterministic world, omniscient private insurance companies would never lose money on a customer, as premiums could be precisely the cost said customer will produce over their contract, plus a small markup; in other words, not providing any insurance, except possibly against inflation.
Keeping this in mind it seems logical that it's in the best interest of society at large to keep insurance companies mostly in the dark.
At least in the case of health insurance, this darkness is threatened by introducing gene testing aided risk analysis.
The inclusion of factors like genetic predisposition to disease in risk analysis is an incredibly helpful tool to doctors when choosing appropriate treatment and taking precautions, but becomes detrimental when accessed by insurance companies\cite{insurancegenes}.
Legal action to protect against this threat is already being taken, with the USA introducing the Genetic Information Nondiscrimination Act in 2008\cite{insurancegenes}.
It is imperative that sensitive data continues to be stored discretely access is tightly controlled as to not enable exploitation of this knowledge.
Were genetic discrimination to become a common occurrence, would it be likely that the trade-off for a chance at a healthier life would no longer be worth it to certain people, which would lead to unnecessary and preventable suffering.
It is therefore vital that we retain trust in and refine the methodology of genetic testing, while at the same time making sure the sensitive information is stored discretely and genetic discrimination is kept under control.

@@

\subsection{Eugenics}
Eugenics, subject to study under the field of bioethics, has been of concern to humanity since the beginning of civilization. While it literally translates to "good birth", it encompasses all attempts at improvement to the biological nature of the human race as a whole\cite{eugenics}. \newline
History has seen it as the justification for a number of crude methods like infanticide in Sparta and forced sterilization under the Nazi regime, some of which are still carried out around the world today. Yet the tools we are soon to wield at the current pace of development towards controlled genetic manipulation go far beyond that, making a new form of eugenics a viable option towards social good. \newline
For the purpose of eugenics, genetic manipulation and genome analysis of fertilized cells to select which one to implement are almost equivalent, as both manipulate chance in our favor. The exact method is therefore irrelevant to the discussion in this chapter.
In itself maximizing potential for generations to come, whether it be with improved intelligence, longer and healthier lives or stress tolerance seems like a noble goal, however one must not belittle possible consequences.

>>>>

\subsection{Genetic Engineering}
Human interference in evolution to bend nature to our will goes back at least to 2000 B.C. with strong evidence that the Assyrians artificially pollinated date trees\cite[p.633]{mayr}.
Since then, several methods with the shared goal of reinforcing desirable characteristics and eradicating undesirables ones developed, most dominantly selective breeding, with the recent modification called mutation breeding, the practice of directly exposing organisms, usually seeds, to mutagenic material to positively augment the mutation rate and such the chance of new positive traits being present\cite{mutationbreeding}.
It can be argued that since selective and mutation breeding only ever affect coming generations it can hardly be seen as genetic modification of a single organism and is better labeled as a separate method of eugenics.
From that perspective the first milestone of genetic engineering was the Avery–MacLeod–McCarty experiment in 1944, which proved that DNA was responsible for transformation in virus strains. It built upon Griffith's experiment from 1928, which showed that potent but dead virus strains can still transfer their properties to usually harmless strains, via something called the "transforming principle".
Avery, MacLeod and McCarty proved that the relevant component for this to work was DNA and not like previously assumed specific proteins, since the phenomenon persist when the solution is treated with protein dissolving substance, but seizes to persist when treated with DNA dissolving agents.

The most direct and controlled form of genetic modification is inserting strings of DNA into existing cells to transfer properties from one organism to another, deactivating- or cutting them out to suppress traits.
The most prominent example of this to date is the CRISPR/Cas method, which accomplishes this semi-reliably via the use of bacteria, but currently still struggles with off target effects, i.e. unforeseen changes outside of the intended locality.
While originally discovered in 1987, its use for genetic editing was only demonstrated in 2011 and since then a tremendous amount of research went into improving effectivity, safety and specificity, as well as finding applications in numerous in agriculture and medicine.
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non-modified cells\cite{crispr}.
This can be worked around in three ways.
Modify a sufficient number of cells (most likely infeasible for broad changes), modify in an early stage of development, where only a number of cells exist, or germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

Secondly gene association studies, which aim to map genetic variants to observable traits and come in two varieties. 
Genome-wide association studies (GWAS) hope to uncover any correlation between strings of DNA and phenotype, while specific genetic association or gene linkage studies focus on specific variants of a gene (alleles) and examine whether they frequently cooccur with disease\cite{gwas}.
This endeavour presents numerous challenges stemming from the sheer complexity of the mechanisms at hand.
Part of the reason this differs from an exact science is changes an organisms environment has on its genome. It is estimated as much 98\% of human DNA is noncoding, id est lacks any apparent purpose\cite{noncoding}. Some of it however is not entirely dysfunctional, rather it is dynamically activated and silenced in a process known as epigenetics changes by so called regulatory genes\cite{epigenetics}.
This means the exact same sequence of DNA can express itself differently, making direct conclusions about the significance of a specific snippet fall short of reliability.

In 2003 with the official completion of the Human Genome Project (HGP) this has been accomplished\cite{HGPFAQ}.
The human genome consists of approximately 3.3 billion base pairs, by cross referencing a healthy genome, single point mutations can be effortlessly identified in theory, especially with modern high-throughput analysis tools\cite{rawal}.
Unfortunately in the context of genetics this notion of a healthy genome is vague. Not only does a healthy genome not exist, simply averaging the genome of a few relatively healthy adults would do it no justice. Changes in DNA can also come in the form of deletion or addition of bases rendering traditional similarity metrics useless as an applicable method would have to be alignment free, in other words agnostic to the exact position in sequence, rather it should recognize subsequences or repetitions anywhere they show up and compare the frequency. Through the nature of mutation, the 'average genome' also varies widely regionally, the HGP in particular faced criticism for overrepresenting caucasian DNA, owed to the fact that it was conducted on anonymous donations of DNA in the USA\cite{HGPcritic}.
It should therefore be seen more like a reference frame to possibly detect deviation.
Cataloguing mutations is useful but by itself often provides little insight, as chances are the mutation occurred a noncoding part.
Even when mutation occurs in regulatory sequences or as part of a gene single mutations these usually have little effect.
Its important to keep in mind that with mutations being so common, each cell has a slightly different sequence. The only mutations they share are those that occurred in the germ cells that formed the zygote or in the very early stages afterwards.
Germline mutations occur roughly $\frac{number of base pairs}{2}*10^{-9}$ times per year. In humans that works out to about 1.65 inheritable mutations per year per parent, but when examining the genome of adult humans there is sampling bias at play. Single point mutations with drastic changes usually just render the cells nonviable, consequently an egg being fertilized and developing long enough to be given birth to indicates a mostly intact genome\cite{scally}.
Somatic mutations occur in the order of magnitude of $ 2*\frac{number of cell divisions}{10^8 base pairs}$, which with 30 trillion cells equates to roughly 500 trillion per year\cite{somaticmutation}.
These usually have no or very little effect, are repaired by the cells dna repair mechanisms or kill the cell either by inhibiting vital function or by causing cell suicide\cite{dnarepair}.
The unlikely unlikely case that DNA damage results in neither of these scenarios and the cell instead starts rapidly reproducing is what we call cancer.

@@

Every individual has a unique genetic makeup, their own distinct form of the human genome. 
Furthermore, our DNA remains the same from the first instant of an individual’s existence to his or her last breath\cite{dnaidentity}.
Sociologist Abels describes identity as the answer to the question \textit{who am I} and therefore, identity as a differentiating term\cite{abelsidentity}. Consequently, it is not unjustified to combine both. On the one hand, there is a distinctive combination of DNA sequences that is individual for everyone and consistent for life. 
On the other hand, there is a need for clarification about one's own identity, the will to differentiate oneself from others while remaining consistent within oneself. But what do people hope to gain from their genetic test results?

According to Nordgren, new information can lead to a new self-understanding\cite{nordgren}.
The results of genetic testing allow an insight into irrefutable truths, when executed correctly by qualified personnel. 
In the case of a paternity test, a good example for this would be "I am P’s father". The statement is added as a new aspect of one's own perception and thus as part of one's own identity. Same procedure for the statement "I am an African-American with my roots in Nigeria", based on genetic ancestry tracing.

In a study published in 2018, about 30\% of respondents said they had taken a genetic test to learn how their DNA influences their physical or emotional traits\cite{canceropinion}. 
69 \% of Americans surveyed are very confident that there is a complex genetic code inside our cells that helps determine "who we are"\cite{apgfk}.
Apparently, the link between DNA results and information about one's own characteristics thus seems to have already reached some people. Whether the link was clear from the start or created through targeted marketing, it is probably a combination of both.

\subsection{Genetic Privacy}
Claiming that something partly shared between a society and mostly shared between members of a lineage can be private is somewhat of an oxymoron in itself. Yet there is an intuitive sense that this fingerprint within each of our nuclei belongs to us and deserves protection like any other deeply individual aspect of our identity.
Unlike spatial privacy, which is concerned with bodily integrity as well as protection of places and material objects we consider our own, genetic privacy falls within the category of informational privacy, grouped with medical records, political and religious belief and facts we wish to keep to ourselves.

What we consider to be private differs from person to person so prerequisite to a discussion on this topic is a shared idea of what privacy actually entails\cite{geneticprivacy}. 
While a unified definition does not exist, privacy blends in with respect and comfort.
The need for privacy stems from insecurity, we wish for our things to be left untouched because we are not sure whoever does keeps them intact, as well as our intimate information be kept private for fear of embarassment.
A close friend entering our home to water the plants feels different from an intruder at midnight.
Genetic data is analogous, knowing a trusted doctor will interpret the results and then discard of it would put an end to the vast majority of concerns. 
We wish for informational privacy because we can never be sure where this intangible information about our future will end up, in addition to worries about what future technology might unveil.
A diagnosis about a broken leg stays the same diagnosis no matter how much time passes, but as soon as a genome is sequenced, more and more insights may be derived from the same test with emerging new methods.
This uncertainty feeds into the general anxiety around the topic already fueled by lack of public education.
Coupled with concerns about the sample collection process for example by religious citizens, these worries dictate legal protection of genetic privacy, as it is the sole way of handling such matters a pluralistic and democratically organized society to ensure both peace of mind and limit abuse of sensitive genetic information\cite{geneticprivacy}.

@@

DNA tests are more popular than ever. Due to the decline in the cost of genetic analysis in the last two decades, "direct-to-consumer" genetic screening has become affordable for more and more people\cite{rodney}. 
In 2018, the global direct-to-consumer market was valued at approximately 824.1 million USD and is estimated to grow to 6,364.5 million USD by 2028. 
The rise in the public awareness along with the reduced cost and time required for sequencing owing to the technological advancements is responsible for bolstering the industry demand and aiding in reduction of the global healthcare expenditure\cite{bisresearch}.

\vspace{0.2cm}
\begin{quote}
"Get insights from your DNA, whether it's your ethnicity or personal traits." (Ancestry.com)
\end{quote}
\vspace{0.2cm}
\begin{quote}
"Find out your likelihood of having certain characteristics. See how your DNA affects your hair color, taste preferences and more." (23andme.com)
\end{quote}
\vspace{0.2cm}

Many genetic testing companies appeal in their rhetoric to identity. The previous quotes use words like "characteristics" and "personal traits" to manifest the connection between DNA testing and the customers identity. 
Nordgreen describes this as genetic individualism which is the view that the DNA of an individual gives the individual a unique identity\cite{nordgren}.
This is in accordance with the statements from the previous chapter about DNA test results leading to a new self-understanding.
Furthermore, Nordgren state that modern societies are searching for identity and this rhetoric is a response to that\cite{nordgren}.
In addition to that, Nordgren and Juengst believe that the business models of these companies represent the confluence of three very different currents within contemporary Euro-American culture: the distinctly pre-modern search for a naturalistic understanding of individual identity in a pluralistic world, the thoroughly modern cachet of genomics as a science, and the post-modern emphasis on radical individual self-determination\cite{nordgrenjuengst}. 
The marketing methods of the websites partly explain the growing popularity of DNA tests, but what are the opportunities and risks?

@@

The two biggest market players Ancestry and 23andMe have a combined DNA Database Size of 32+ million\cite{ancestrydatabase}\cite{about23andme}. With this amount of data and due to the sensitivity of the content, it should be handled very responsibly. But how does it look in practice?
Laestadius, Rich \& Auer investigated 30 different direct-to-consumer genetic testing websites for two aspects. 
The first aspect being transparency regarding confidentiality and privacy. As a second aspect, the use of data, professional and government bodies created guidelines to promote transparency among these companies was analysed. 
The authors concluded that although most companies met guidelines related to transparency regarding security protocols, storage procedures, and third-party disclosures, only few met guidelines regarding sharing risks from data disclosures. 
Additionally, few companies disclosed how long data would be kept for services or research. 
Use of data for research was frequently mentioned only in privacy policies and terms of service documents, and only two-thirds of companies required an additional consent to use consumer data for health-related research. 
This analysis shows that companies do not consistently meet international transparency guidelines related to confidentiality, privacy, and secondary use of data and are therefore a risk for their customers. 
The authors appeal to clearly inform consumers about specific third parties that will be given access to their data, whether they can have their samples and data deleted at will, and whether their data will be used for anything other than the services they purchased\cite{companydata}.

@@

The controversy around privacy versus surveillance applies to genome analysis as much as to any sensitive technology. 
Germanys BKA describes the evolution of the usage of DNA analysis as going from a supplement to the "classic" identification method of fingerprints to an equal and self-evident means of securing evidence\cite{bkapres}. 
A common claim of advocates for DNA collection is that there is no difference between DNA and fingerprinting in terms of its invasion of an individual's privacy. 
But while fingerprints are a unique identifier, DNA identifies networks of people\cite{cwduster}. DNA data can reveal information about third parties that they have not consented to.
 In the case of the Golden state killer, Criminal DNA databases produced no hits but a match to a distant relative on a genealogy website has led to the identity and therefore the arrest of the perpetrator\cite{goldenstate}. 
In this incident, the misuse of DNA profiles on genealogy websites has led to the prevention of further crimes and therefore has added value for society.

@@

\subsection{Paternity}
After testing for genetic relatedness, non-legal and legal paternity tests are offered most often by direct-to-consumer genetic companies\cite{dnastatista}.
Genetic tests can be helpful in establishing evidence for the parenthood of a person for a case like child custody and support. 
The results of genetic test can also be used as a support for placing a parent’s name on the birth certificate of a child or for settling disputes in child custody laws. However, whether the sampling is done with the consent of all parties tested is not the responsibility of the genetic websites.

@@

\subsection{Medical Sector}
As preventing disease is usually preferable over treating it, a substantial amount of medicine focuses on early detection and prevention. Especially for the class of mendelian disorders, which are characterized by being linkable to a single gene defect, genetic testing has proven to make diagnosis easy and reliable \cite{rawal}.
However while a grand number of such diseases exist, they are usually rare so testing might only pay off if the defect is known to run in the family \cite{publicopinion}.

Much bigger potential lies in preventing common disease with either a genetic component or whose causes are spread over several genes and are therefore  harder to pinpoint. This uncertainty neccessarily introduces stochastics and the method known as genetic risk assessment \cite{gwas}.
Advances here have sprouted the field of pharmagenomics, which utilizes genetic information to practice precision medicine.
Just like ethnicity, genetic profile has been shown to affect drug response, thus using this knowledge to personalize prescriptions can increase effectiveness and lower cost and side effects \cite{costsaving}.
Predictive genetic testing does not come without concerns from both experts and the public, the most prominent ones will be discussed in the following:

\subsubsection{Patient Autonomy}
The primary role of any new medical instrument should be to serve, therefore patient autonomy and informed consent are non-negotiable, which necessitates  public education.
Broad, as well as targeted at populations most susceptible to specific genetic conditions.
Equally as important is access to testing facilities themselves as well as short wait times for both test and treatment\cite{publicopinion}.

\subsubsection{Genetic Discrimation}
As will be discussed in the following chapter, subpar treatment grounded in genome analysis results is no better than discrimination on the basis of sex, race, age or religion.

\subsubsection{Commercial Use}
Profit as the primary incentive manages to corrupt even the most honourable endeavour.
Commercial gene testing and therapy raises the same questions about motive and ethical obligation to help that plagued the medical ethics for millennia.
Testing costs being inflated because of patents can hardly be in accordance with public interest, which sparks discussion about the next point\cite{publicopinion}.

\subsubsection{Use of Public Resources}
If private patents are to be avoided, public resources have to be used towards research in the applicable area.
Distributing those in a fair way however proves difficult. Developing tests is expensive so some conditions need to be prioritized, which in turn requires weighing of different factors. The initial cost of development is just as important as benefit it brings. This is comprised of the reliability of the result, as well as therapy options versus their respective cost. An unreliable result is equally useless from a public health perspective as knowledge about a genetic death sentence.
Additionally as many tests would only ever affect a small number of people, some argue that the money should either be used for more general genetic tests or towards programs that target public health in a broader sense like addressing diet\cite{publicopinion}.

\subsubsection{Patient Responsibilities}
While a general right to data privacy in medicine makes sense for traditional diagnostics, one might argue that since DNA test results have immediate consequences for relatives theres at least in part a responsibility to inform said relatives of any risks they themselves might face, comparable to infectious disease or STDs\cite{publicopinion}.
This apparent obligation to warn stands in direct conflict with the testers pledge of silence and the testees right to privacy, a conflict that could in theory be resolved with institutions for anonymous testing and subsequent information of concerned parties analogous to HIV testing.

\subsubsection{Psychological Risks}
The psychological weight associated with carrying near certain information about the future is not to be underestimated in the context of genetic testing either.
Just as a cancer diagnosis, discovering to be at high risk of developing potentially lethal disease can be devastating, especially if treatment options are limited\cite{publicopinion}.
Depending on the evolution of public debate on the topic, a scenario in which gene therapy is frowned upon in a similar manner to termination of pregnancy by groups within society is also imaginable.
This would add additional stress and force people at risk for genetic disease to make a choice between social acceptability and their own health.

\subsection{Insurance}
Insurance as a business walks a fine line between solidarity and preying on the vulnerable. Risk and how we manage it as a society is studied under the name of insurance ethics, but to date the literature is sparse.
While public insurance in practice is a form of wealth redistribution, private insurance is in essence paying for peace of mind\cite{insuranceethics}.
Looking at the probability theory behind it, we expect to pay more than the 
However this social safety net only functions as long as there is uncertainty, in fact it functions better the more uncertainty there is.
This becomes obvious in consideration of the idea that clarity about the future replaces the insurance sector.
In a fully deterministic world, omniscient private insurance companies would never lose money on a customer, as premiums could be precisely the cost said customer will produce over their contract, plus a small markup; in other words, not providing any insurance, except possibly against inflation.
Keeping this in mind it seems logical that it's in the best interest of society at large to keep insurance companies mostly in the dark.
At least in the case of health insurance, this darkness is threatened by introducing gene testing aided risk analysis.
The inclusion of factors like genetic predisposition to disease in risk analysis is an incredibly helpful tool to doctors when choosing appropriate treatment and taking precautions, but becomes detrimental when accessed by insurance companies\cite{insurancegenes}.
Legal action to protect against this threat is already being taken, with the USA introducing the Genetic Information Nondiscrimination Act in 2008\cite{insurancegenes}.

It is imperative that sensitive data continues to be stored discretely access is tightly controlled as to not enable exploitation of this knowledge.
Were genetic discrimination to become a common occurrence, would it be likely that the trade-off for a chance at a healthier life would no longer be worth it to certain people, which would lead to unnecessary and preventable suffering.
It is therefore vital that we retain trust in and refine the methodology of genetic testing, while at the same time making sure the sensitive information is stored discretely and genetic discrimination is kept under control.

@@

\subsection{Eugenics}
Eugenics, subject to study under the field of bioethics, has been of concern to humanity since the beginning of civilization. While it literally translates to "good birth", it encompasses all attempts at improvement to the biological nature of the human race as a whole\cite{eugenics}.

History has seen it as the justification for a number of crude methods like infanticide in Sparta and forced sterilization under the Nazi regime, some of which are still carried out around the world today. Yet the tools we are soon to wield at the current pace of development towards controlled genetic manipulation go far beyond that, making a new form of eugenics a viable option towards social good. \newline
For the purpose of eugenics, genetic manipulation and genome analysis of fertilized cells to select which one to implement are almost equivalent, as both manipulate chance in our favor. The exact method is therefore irrelevant to the discussion in this chapter.
In itself maximizing potential for generations to come, whether it be with improved intelligence, longer and healthier lives or stress tolerance seems like a noble goal, however one must not belittle possible consequences.
